Data-based completion of grey-box models with neural networks

### 2020-10-15 systemID_NNs.tex
will work slightly to modify the plant cost curve.

in which $x_g$ are the grey-box states, which are some subset 
of the plant states (denote $x_g = Gx$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k:k-N} &= [y(k)', y(k-\Delta)', ..., y(k-N\Delta)']' \\
  \mathbf{u}_{k-1:k-N} &= [u(k-1)', u(k-2)', ..., u(k-N\Delta)']'
\end{align*}

in which $\Delta$ is the sample 
time between two consecutive
prediction instants of the desired discrete time 
model to be used in MPC. There exists a $\phi(\cdot)$ 
and a number of delay 
time steps $N$, such that the plant state
at a given time can be written as
$x(k) = \phi(\mathbf{y}_{k:k-N}, \mathbf{u}_{k-1:k-N})$.

We augment the grey-box model using 
neural networks as follows -
\begin{align*}
\dot x_g &= f_g(x_g, u) + 
f_{NN}(\mathbf{y}_{k:k-N}, \mathbf{u}_{k:k-N})\\
y &= h_g(x_g) + h_{NN}(\mathbf{y}_{k:k-N}, \mathbf{u}_{k-1:k-N})
\end{align*}
in which $f_{NN}(\cdot)$ and $h_{NN}(\cdot)$
are standard feedforward neural networks. We add the current 
control $u(k)$ as an input to the network $f_{NN}(\cdot)$ because 
there may be a mismatch in the actuator to the grey-box dynamics
as well.

It is expected that the functions $f_{NN}(\cdot)$ 
and $h_{NN}(\cdot)$ approximate
the following differences between the true grey-box state dynamics
and true measurement function with what is modeled.
\begin{align*}
  f_{NN}(\mathbf{y}_{k:k-N}, \mathbf{u}_{k:k-N}) &\approx 
      Gf_p(\phi(\mathbf{y}_{k:k-N}, \mathbf{u}_{k-1:k-N}), u) - f_g(x_g, u) = 
      Gf_p(x, u) - f_g(x_g, u) \\
      h_{NN}(\mathbf{y}_{k:k-N}, \mathbf{u}_{k-1:k-N}) &\approx 
      h_p(\phi(\mathbf{y}_{k:k-N}, \mathbf{u}_{k-1:k-N})) - h_g(x_g) = 
      h_p(x) - h_g(x_g) 
\end{align*}

@@

\begin{align*}
  k_1 &= f_g(x_g(k), u(k)) + f_{NN}(\mathbf{y}_{k:k-N}, \mathbf{u}_{k:k-N}) \\
  k_2 &= f_g(x_g(k) + (\Delta/2)k_1, u(k)) + 
         f_{NN}(\mathbf{y}_{k+0.5:k-N+0.5}, 
                \mathbf{u}_{k+0.5:k-N+0.5})\\
  k_3 &= f_g(x_g(k) + (\Delta/2)k_2, u(k)) + 
                f_{NN}(\mathbf{y}_{k+0.5:k-N+0.5}, 
                      \mathbf{y}_{k+0.5:k-N+0.5})\\
  k_4 &= f_g(x_g(k) + \Delta k_3, u(k)) + 
                       f_{NN}(\mathbf{y}_{k+1:k-N+1}, 
                              \mathbf{u}_{k+1:k-N+1})\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) + h_{NN}(\mathbf{y}_{k+\Delta:k-N+\Delta}, 
                                     \mathbf{u}_{k-1+\Delta:k-N+\Delta})
\end{align*}
in which the sequences $\mathbf{y}_{k+0.5:k-N+0.5}$ 
and $\mathbf{u}_{k+0.5:k-N+0.5}$ are defined as 
\begin{align*}
  \mathbf{y}_{k+0.5:k-N+0.5} &= 
    [y(k + \Delta/2)', y(k-\Delta/2)', ..., y(k-N\Delta + \Delta/2)']' \\
  \mathbf{u}_{k+0.5:k-N+0.5} &= 
  [u(k + \Delta/2)', u(k-\Delta/2)', ..., u(k-N\Delta + \Delta/2)']' 
\end{align*}
We assume a zero order hold on the control inputs between two consecutive 
MPC execution instants, such that $u(k + \Delta/2) = u(k)$. To get the 
measurements between two sample instants ($y(k + \Delta/2)$) for 
the RK4 predictions, a linear interpolation can be used 
or we can assume that additional 
measurements from the plant are available 
between two MPC execution instants as well.

The following optimization problem is solved to compute 
the weights in the neural networks 
\begin{align*}
 \underset{W_{fi}, b_{fi}, W_{hi}, b_{hi}}{\textnormal{min}} \sum_{i=0}^{N_s-1} 
 \dfrac{1}{2}(\hat{y}(i) - y(i))^2
\end{align*}
in which ($W_{fi}, b_{fi}, W_{hi}, b_{hi}$) are the weights in the 
two neural networks ($f_{NN}(\cdot)$ and $h_{NN}(\cdot)$), 
$\hat{y}$ are predictions of the augmented model, and $y$
are the measurements from the plant.
Note that this optimization requires an estimate 
of the initial grey-box state. We assume that 
before the open-loop system identification experiment 
performed for data collection, 
the plant is being operated at some 
steady-state such that input to output
responses of the plant and grey-box model are aligned with 
the use of an integrating disturbance model. We can use 
the grey-box state at that steady state as the initial condition 
for the above multi-step-ahead prediction-error
optimization.

\subsection{Simple example}

@@

\textbf{Plant --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
These ODEs represent the dynamics of the reaction system
$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
height and temperature. The control input is $u = C_{A0}$, 
and the measurements are $y = [C_A, C_B]'$. The plant model 
is observable with these choices of the measurements and dynamics.

@@

>>>>

in which $x_g$ are the grey-box states, which are some subset 
of the plant states (denote $x_g = Gx$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k+1-N_p:k} &= [y(k)', y(k-\Delta)', ..., y(k-(N_p-1)\Delta)']' \\
  \mathbf{u}_{k+1-N_p:k-1} &= [u(k-\Delta)', u(k-2\Delta)', 
                               ..., u(k-(N_p-1)\Delta)']'
\end{align*}

in which $N_p$ is the number of 
past measurements and
$\Delta$ is the sample time between two consecutive
measurements. There exists a $\phi(\cdot)$ 
and a number of delay 
time steps $N_p$, such that the plant state
at a given time can be written as
$x(k) = \phi(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})$.
For linear plants, this function $\phi(\cdot)$
is known analytically in terms of the plant dynamic
matrices.

We augment the grey-box model using 
neural networks as follows -
\begin{align*}
\dot x_g &= f_g(x_g, u) + 
f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k})\\
y &= h_g(x_g) + h_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})
\end{align*}
in which $f_{NN}(\cdot)$ and $h_{NN}(\cdot)$
are standard feedforward neural networks. We add the current 
control $u(k)$ as an input to the network $f_{NN}(\cdot)$ because 
there may be a mismatch in the actuator to the grey-box dynamics
as well.

It is expected that the functions $f_{NN}(\cdot)$ 
and $h_{NN}(\cdot)$ approximate
the following differences between the true grey-box state dynamics
and true measurement function with what is modeled.
\begin{align*}
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k}) &\approx 
      Gf_p(x, u) - f_g(Gx, u) \\
      h_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) &\approx
      h_p(x) - h_g(Gx) 
\end{align*}  
in which $x(k) = \phi(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})$
is an unknown function of past measurements and inputs.

@@

\begin{align*}
  k_1 &= f_g(x_g(k), u(k)) + 
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k}) \\
  k_2 &= f_g(x_g(k) + (\Delta/2)k_1, u(k)) + 
  f_{NN}(\mathbf{y}_{k+1.5-N_p:k+0.5}, \mathbf{u}_{k+1.5-N_p:k+0.5})\\
  k_3 &= f_g(x_g(k) + (\Delta/2)k_2, u(k)) + 
  f_{NN}(\mathbf{y}_{k+1.5-N_p:k+0.5}, \mathbf{u}_{k+1.5-N_p:k+0.5})\\
  k_4 &= f_g(x_g(k) + \Delta k_3, u(k)) + 
  f_{NN}(\mathbf{y}_{k+2-N_p:k+1}, \mathbf{u}_{k+2-N_p:k+1})\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) + 
  h_{NN}(\mathbf{y}_{k+2-N_p:k+1}, \mathbf{u}_{k+2-N_p:k+1})
\end{align*}
in which the sequences $\mathbf{y}_{k+1.5-N_p:k+0.5}$ 
and $\mathbf{u}_{k+1.5-N_p:k+0.5}$ are defined as 
\begin{align*}
  \mathbf{y}_{k+1.5-N_p:k+0.5} &= 
    [y(k + \Delta/2)', y(k-\Delta/2)', ..., y(k-(N_p-1)\Delta + \Delta/2)']' \\
    \mathbf{u}_{k+1.5-N_p:k+0.5} &= 
  [u(k + \Delta/2)', u(k-\Delta/2)', ..., u(k-(N_p-1)\Delta + \Delta/2)']' 
\end{align*}
We assume a zero order hold on the control inputs between two consecutive 
MPC execution instants, such that $u(k + \Delta/2) = u(k)$. To get the 
measurements between two sample instants (for e.g., $y(k + \Delta/2)$) for 
the RK4 predictions, linear interpolation is used.

The following optimization problem is solved to compute 
the weights in the neural networks 
\begin{align*}
 \underset{W_{fi}, b_{fi}, W_{hi}, b_{hi}}{\textnormal{min}} \sum_{k=0}^{N_s-1} 
 \dfrac{1}{2}(\hat{y}(k) - y(k))^2
\end{align*}
in which ($W_{fi}, b_{fi}, W_{hi}, b_{hi}$) are the weights in the 
two neural networks ($f_{NN}(\cdot)$ and $h_{NN}(\cdot)$), 
$\hat{y}$ are the predictions of the augmented
grey-box + neural network model, and $y$
are the measurements from the plant.
Note that this optimization requires an estimate 
of the initial grey-box state. We assume that 
before the open-loop system identification experiment 
performed for data collection, 
the plant is being operated at some 
steady-state such that input to output
responses of the plant and grey-box model are aligned with 
the use of an integrating disturbance model. We can use 
the grey-box state at that steady state as the initial condition 
for the above multi-step-ahead prediction-error
optimization.

\subsection{Linear plant}

@@

\textbf{Plant --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
These ODEs represent the dynamics of the reaction system
$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
height and temperature. The control input is $u = C_{A0}$,
the states are $x = [C_A, C_B, C_C]'$, 
and the measurements are $y = [C_A, C_B]'$. 
Assume the following 
parameters for this model, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.3 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.2 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
The plant model 
is linear ($dx/dt = Ax + Bu$) and 
observable with the choices of 
the parameters and measurements.

@@

The current state $x(k)$ can be constructed using the history of 
past measurements and inputs as 
\begin{equation*}
  x(k) = A^{N_p-1}_d
  \mathcal{O}^{\dagger}(\mathbf{y}_{k+1-N_p:k} - 
                        \mathcal{B}\mathbf{u}_{k+1-N_p:k-1}) + \\ 
      \mathcal{B}_{nN_p-n+1:nN_p}\mathbf{u}_{k+1-N_p:k-1}
\end{equation*}

in which $A_d$ is the plant state transition matrix in discrete-time, 
$\mathcal{O}$ is the observability matrix, 
and $\mathcal{B}$ is a combination of the 
discrete time matrices ($A_d$ and $B_d$) such that it 
denotes the response of a vector of inputs to states, 
and $\mathcal{B}_{nN_p-n+1:nN_p}$ denotes the last $n$
rows of $\mathcal{B}$.

Since the measurements are a simple linear function 
of the grey-box and plant states, we use only one 
neural network to 
augment the grey-box dynamics. 
The expected function 
to be approximated with the network is
\begin{align*}
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) \approx 
  \begin{bmatrix}
  0 & 0 & 0 \\  
  0 & -k_2 & k_3 
  \end{bmatrix}x(k)
\end{align*}

In this example, the network just approximates a linear function.
The figure below shows validation performances of the grey-box 
and augmented hybrid models. The structure 
of the network is restricted to just a matrix and
the augmented model was trained on 120 samples.

\begin{figure}[!h]
  \centering
  \includegraphics[page=2, width=\textwidth]{tworeac_plots_lin.pdf}
\end{figure}

\subsection{Nonlinear plant}
The plant is made nonlinear by changing the side reaction 
to a third order reaction ($A \rightarrow B $ and  $3B \rightleftharpoons C$),
and is now simulated with the following ODEs
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

The grey-box model is kept the same. The parameters 
in the model are modified to 
$k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.05 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.1 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
An augmented model with a neural
network architecture [8, 4, 2] is trained on 180 samples. The 
tanh activation is used in the network. The figure below shows 
validation performances of the grey-box and augmented 
hybrid models.

\begin{figure}[!h]
  \centering
  \includegraphics[page=2, width=\textwidth]{tworeac_plots_nonlin.pdf}
\end{figure}
